---
title: "Multiple Regression Analysis on Lo 30 Portfolio Returns: A Fama–French 3- and 5-Factor Approach"
author: | 
  | Wei Chieh Chen
  | Syracuse University
date: "2025-04-01"
output: 
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(data.table)
# Load required libraries
library(readr)
library(dplyr)
library(lubridate)
library(car)
library(nortest)
# Load the lmtest package for the Breusch-Pagan test
library(lmtest)
library(broom)

#####################################################
# Create a temporary file for the ZIP archive
temp_zip <- tempfile(fileext = ".zip")

# Download the ZIP file
download.file("https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_Factors_weekly_CSV.zip", temp_zip)

# List the files in the ZIP archive and assume the CSV is the first file
files_in_zip <- unzip(temp_zip, list = TRUE)$Name
csv_file_name <- files_in_zip[1]

# Extract the CSV file to a temporary directory
extracted_path <- tempdir()
unzip(temp_zip, files = csv_file_name, exdir = extracted_path)

# Construct the full path to the extracted CSV file
csv_path <- file.path(extracted_path, csv_file_name)

# Read the CSV file using fread
ff3w_data <- fread(csv_path)
head(ff3w_data)

colnames(ff3w_data)[1] <- "Date"

# Optionally, remove the temporary ZIP file
unlink(temp_zip)

######################################################
# Create a temporary file for the ZIP archive
temp_zip <- tempfile(fileext = ".zip")

# Download the ZIP file
download.file("https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/F-F_Research_Data_5_Factors_2x3_daily_CSV.zip", temp_zip)

# List the files in the ZIP archive and assume the CSV is the first file
files_in_zip <- unzip(temp_zip, list = TRUE)$Name
csv_file_name <- files_in_zip[1]

# Extract the CSV file to a temporary directory
extracted_path <- tempdir()
unzip(temp_zip, files = csv_file_name, exdir = extracted_path)

# Construct the full path to the extracted CSV file
csv_path <- file.path(extracted_path, csv_file_name)

# Read the CSV file using fread
ff5d_data <- fread(csv_path)
head(ff5d_data)

colnames(ff5d_data)[1] <- "Date" 

# Optionally, remove the temporary ZIP file
unlink(temp_zip)

######################################################
# Create a temporary file for the ZIP archive
temp_zip <- tempfile(fileext = ".zip")

# Download the ZIP file
download.file("https://mba.tuck.dartmouth.edu/pages/faculty/ken.french/ftp/Portfolios_Formed_on_ME_Daily_CSV.zip", temp_zip)

# List the files in the ZIP archive and assume the CSV is the first file
files_in_zip <- unzip(temp_zip, list = TRUE)$Name
csv_file_name <- files_in_zip[1]

# Extract the CSV file to a temporary directory
extracted_path <- tempdir()
unzip(temp_zip, files = csv_file_name, exdir = extracted_path)

# Construct the full path to the extracted CSV file
csv_path <- file.path(extracted_path, csv_file_name)

# Read the CSV file using fread
pfosd_data <- fread(csv_path, header = TRUE)
head(pfosd_data)

# Optionally, remove the temporary ZIP file
unlink(temp_zip)
colnames(pfosd_data)[1] <- "Date" 
head(pfosd_data)

# Process Fama-French 3-Factor data:
ff3 <- ff3w_data %>%
  mutate(Date = ymd(as.character(Date))) %>%    # Convert Date column to Date format
  filter(Date >= ymd("1963-07-01"))              # Restrict data from July 1, 1963 onward

# Process Fama-French 5-Factor data:
ff5 <- ff5d_data %>%
  mutate(Date = ymd(as.character(Date))) %>%  # Convert Date column to Date format
  filter(Date >= ymd("1963-07-01"))           # Restrict data from July 1, 1963 onward

# Process the portfolios data:
portfolios <- pfosd_data %>%
  mutate(Date = ymd(as.character(Date))) %>%  
  filter(Date >= ymd("1963-07-01"))

# Adjust column names: Use make.names() to handle spaces and special characters
colnames(portfolios) <- make.names(colnames(portfolios))
# Assume that "Lo 30" becomes "Lo.30" and rename it to "Lo30" for convenience.
portfolios <- portfolios %>%
  rename(Lo30 = Lo.30)

# Merge portfolios with Fama-French 3-Factor data:
data_3f <- inner_join(portfolios, ff3, by = "Date")

# Merge portfolios with Fama-French 5-Factor data:
data_5f <- inner_join(portfolios, ff5, by = "Date")

# Ensure numeric conversion for model variables
data_3f$`Mkt-RF` <- as.numeric(as.character(data_3f$`Mkt-RF`))
data_3f$SMB <- as.numeric(as.character(data_3f$SMB))
data_3f$HML <- as.numeric(as.character(data_3f$HML))
data_3f$Lo30 <- as.numeric(as.character(data_3f$Lo30))

# Rename column for convenience
colnames(data_3f)[which(names(data_3f) == "Mkt-RF")] <- "MKTRF"

# Fit the model: Lo30 ~ MKTRF + SMB + HML
model_3f <- lm(Lo30 ~ MKTRF + SMB + HML, data = data_3f)
summary(model_3f)

data_5f$MKTRF  <- as.numeric(as.character(data_5f$MKTRF))
data_5f$SMB    <- as.numeric(as.character(data_5f$SMB))
data_5f$HML    <- as.numeric(as.character(data_5f$HML))
data_5f$RMW    <- as.numeric(as.character(data_5f$RMW))
data_5f$CMA    <- as.numeric(as.character(data_5f$CMA))
data_5f$Lo30   <- as.numeric(as.character(data_5f$Lo30))

# 5-Factor Regression: Lo 30 ~ Mkt-RF + SMB + HML + RMW + CMA
colnames(data_5f)[21] <- "MKTRF"
model_5f <- lm(Lo30 ~ MKTRF + SMB + HML + RMW + CMA, data = data_5f)
summary(model_5f)
```

## Introduction

\textbf{Overview:}
This project demonstrates how to perform two multiple linear regression analyses using the “Lo 30” portfolio returns as the response variable. Two sets of factor datasets are used:

- Fama–French 3-Factor Model: Market excess return (Mkt-RF), SMB, and HML.

- Fama–French 5-Factor Model: The above factors plus RMW and CMA.

Both models are estimated for data from July 1, 1963, onward.

## Data Import
\textbf{Summary of Steps:}

- Downloaded ZIP archives for three datasets from online sources:

  - Fama-French 3-Factor Data (Weekly)

  - Fama-French 5-Factor Data (Daily)

  - Portfolios Data (Daily)

- Created temporary ZIP files and extracted the first CSV file from each archive.

- Read the CSV files using fread() and previewed the data.

- Renamed the first column to "Date" for consistency.

- Optionally removed temporary ZIP files to clean up.

## Data Preparation & Merging the Datasets

\textbf{Data Preparation Summary:}

- Convert "Date" columns to proper Date objects using ymd().

- Filter each dataset to include only records from July 1, 1963 onward.

- Adjust column names in the portfolios dataset with make.names() and rename "Lo 30" (i.e., "Lo.30") to "Lo30".

\textbf{Merging Summary:}

- Merge the portfolios data with the Fama-French 3-Factor data by the "Date" variable.

- Merge the portfolios data with the Fama-French 5-Factor data similarly to ensure matching dates for regression analysis.

## Display summary of merged data (example)
<div style="font-size: 0.8em;">
```{r, echo=TRUE}
summary(data_3f[, c(21:23)])
```

## Regression Analysis – Fama-French 3-Factor Model

- Fit the 3-Factor regression model and display the summary.

```{r reg1, echo=FALSE, message=FALSE, warning=FALSE}
# Tidy the model output and select the coefficient table columns
coef_table <- tidy(model_3f)[,1:5]
# Display the table with a caption
knitr::kable(coef_table, caption = "Coefficient Table for Fama-French 3-Factor Model")
```


## Regression Analysis – Fama-French 3-Factor Model
\textbf{Interpretation:}

- The intercept, MKTRF, and SMB are statistically significant.

- HML is not significant.

- Approximately 21% of the variance in Lo30 is explained.



## Regression Analysis – Fama-French 5-Factor Model

```{r reg2, echo=FALSE, message=FALSE, warning=FALSE}
# Tidy the model output and select the coefficient table columns
coef_table <- tidy(model_5f)[,1:5]
# Display the table with a caption
knitr::kable(coef_table, caption = "Coefficient Table for Fama-French 5-Factor Model")
```

- All predictors are statistically significant.

- The model explains over 98% of the variance in Lo30.

## Regression Diagnostics – Fama-French Models

The following plots assess the regression assumptions (linearity, homoscedasticity, normality, and influential observations) for both models.  
The top row shows the diagnostics for the 3-Factor model, while the bottom row shows those for the 5-Factor model.

```{r diag-both, fig.height=5, fig.width=10, echo=FALSE}
# Set up the plotting area for 8 plots (2 rows x 4 columns)
par(mfrow = c(2, 4))

# Diagnostic plots for the 3-Factor model
plot(model_3f, which = 1:4)

# Diagnostic plots for the 5-Factor model
plot(model_5f, which = 1:4)
```



## Conclusion & Next Steps
\textbf{Key Findings:}

- The 3-Factor model shows moderate explanatory power.

- The 5-Factor model fits exceptionally well but shows some departures from classical assumptions.

\textbf{Diagnostics Summary:}

- Both models have issues with non-normality and heteroscedasticity.

- Future work should consider robust standard errors or alternative estimation techniques.

\textbf{Next Steps:}

- Explore model refinements.

- Validate findings with out-of-sample tests.